\section{Linear operations on sorted ranges}

\subsection{includes}

The first linear algorithm we will talk about is std::includes. This algorithm determines whether one range is a sub-range of another. Since we are working on sorted ranges, std::includes runs in linear time.

\cppversions{\texttt{includes}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{(input\_range, input\_range)}}{\texttt{forward\_range, forward\_range)}}{\texttt{operator<}}{\texttt{strict\_weak\_ordering}}

Here we check whether the input text contains all the lowercase English letters.

\begin{box-note}
\begin{cppcode}
std::vector<char> letters('z'-'a'+1,'\0');
std::iota(letters.begin(), letters.end(), 'a');

std::string input = "the quick brown fox jumps over the lazy dog";
std::ranges::sort(input);

assert(std::ranges::includes(input, letters));
\end{cppcode}
\end{box-note}

First, we generate the list of English letters programmatically using another algorithm std::iota (line 2). The iota algorithm generates consecutively increasing values to fill the given range. Because of this, we preallocate the vector to 26 elements (line 1).

\subsection{merge, inplace\_merge}

Another operation that is feasible in linear time is merging two sorted ranges.

\cppversions{\texttt{merge}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{(input\_range, input\_range)}}{\texttt{(forward\_range, forward\_range)}}{\texttt{operator<}}{\texttt{strict\_weak\_ordering}}

The result of the merge operation is stored using the provided output iterator. Note that the output range is not permitted to overlap with either of the input ranges.

The merge operation is stable. Equal elements from the first range will be ordered before equal elements from the second range.

\begin{box-note}
\begin{cppcode}
struct LabeledValue {
    int value;
    std::string label;
};

std::vector<LabeledValue> data1{{1, "first"}, {2, "first"}, {3, "first"}};
std::vector<LabeledValue> data2{{0, "second"}, {2, "second"}, {4, "second"}};

std::vector<LabeledValue> result;
std::ranges::merge(data1, data2, std::back_inserter(result),
  [](const auto& left, const auto& right) { return left.value < right.value; });
// result == {0, second}, {1, first}, {2, first}, {2, second}, {3, first}, {4, second}
\end{cppcode}
\end{box-note}

The parallel version requires the output to be a forward range (represented by a forward\_iterator). Therefore, we cannot use wrappers like std::back\_inserter and must preallocate the output range to sufficient capacity.

\begin{box-note}
\begin{cppcode}
std::vector<int> data1{1, 2, 3, 4, 5, 6};
std::vector<int> data2{3, 4, 5, 6, 7, 8};

std::vector<int> out(data1.size()+data2.size(), 0);
std::merge(std::execution::par_unseq,
    data1.begin(), data1.end(),
    data2.begin(), data2.end(),
    out.begin());
\end{cppcode}
\end{box-note}

\cppversions{\texttt{inplace\_merge}}{\CC98}{N/A}{\CC17}{\CC20}

\constraints{\texttt{(bidirectional\_range, bidirectional\_iterator)}}{}{\texttt{operator<}}{\texttt{inplace\_merge}}

Because merge forbids the input and output ranges from overlapping, we have an alternative inplace\_merge that serves this use case.

\begin{box-note}
\begin{cppcode}
std::vector<int> range{1, 3, 5, 2, 4, 6};
std::inplace_merge(range.begin(), range.begin()+3, range.end());
// range == { 1, 2, 3, 4, 5, 6 }
\end{cppcode}
\end{box-note}

\subsection{unique, unique\_copy}

The std::unique algorithm removes consecutive duplicate values. The typical use case is in conjunction with a sorted range. However, this is not a requirement of std::unique.

\cppversions{\texttt{unique}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{forward\_range}}{}{\texttt{operator==}}{\texttt{binary\_predicate}}

Because unique works in-place and cannot resize the underlying range, it leaves the end of the range with unspecified values and returns an iterator to the beginning of this sub-range (or the sub-range in case of the range version).

\begin{box-note}
\begin{cppcode}
std::vector<int> data{1, 1, 2, 2, 3, 4, 5, 6, 6, 6};
auto it = std::unique(data.begin(), data.end());
// Range version: auto it = std::ranges::unique(data).begin();

// data == {1, 2, 3, 4, 5, 6, unspec, unspec, unspec, unspec}
data.resize(std::distance(data.begin(), it));
// data == {1, 2, 3, 4, 5, 6}
\end{cppcode}
\end{box-note}

\cppversions{\texttt{unique\_copy}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{input\_range -> output\_iterator}}{\texttt{forward\_range -> forward\_iterator}}{\texttt{operator==}}{\texttt{binary\_predicate}}

The copy version of unique instead outputs the unique values to an output range represented by an iterator.

\begin{box-note}
\begin{cppcode}
std::vector<int> data{1, 1, 2, 2, 3, 4, 5, 6, 6, 6};
std::vector<int> out;
std::ranges::unique_copy(data, std::back_inserter(out));
// out == {1, 2, 3, 4, 5, 6}
\end{cppcode}
\end{box-note}