\section{Copy and move}

\subsection{copy, move, copy\_backwards, move\_backwards}

When copying/moving a range, we need to avoid overwriting the yet to be copied/moved elements from the source range. The standard offers two directions of copying to prevent this issue. We will start with the forward version:

\cppversions{\texttt{copy}}{\CC98}{\CC20}{\CC17}{\CC20}
\cppversions{\texttt{move}}{\CC11}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{input\_range -> output\_iterator}}{\texttt{forward\_range -> forward\_iterator}}{}{}

The output iterator is not permitted to be within [first, last) (of the input range). Consequently, only the tail of the output range can overlap with the input range.

\begin{box-note}
\begin{cppcode}
std::vector<std::string> data{ "a", "b", "c", "d", "e", "f"};

std::copy(data.begin(), data.begin()+3, data.begin()+3);
// data = { "a", "b", "c", "a", "b", "c" }

// Overlapping case:
std::copy(std::next(data.begin()), data.end(), data.begin());
// data = { "b", "c", "a", "b", "c", "c" }
\end{cppcode}
\end{box-note}

Move operates identically, except it casts each element to an rvalue before the assignment, turning copies into moves.

\begin{box-note}
\begin{cppcode}
std::vector<std::string> data{ "a", "b", "c", "d", "e", "f"};

std::move(data.begin(), data.begin()+3, data.begin()+3);
// data = { ?, ?, ?, "a", "b", "c" }
\end{cppcode}
\end{box-note}

Significantly, whether std::move will move depends on the underlying element type. If the underlying type is copy-only, std::move will behave identically to std::copy.

\begin{box-note}
\begin{cppcode}
struct CopyOnly {
    CopyOnly() = default;
    CopyOnly(const CopyOnly&) = default;
    CopyOnly& operator=(const CopyOnly&) { 
      std::cout << "Copy assignment.\n";
      return *this;
    };
};

std::vector<CopyOnly> test(6);

std::move(test.begin(), test.begin()+3, test.begin()+3);
// 3x Copy assignment
\end{cppcode}
\end{box-note}

We can also emulate std::move by using the previously mentioned move\_iterator adapter.

\begin{box-note}
\begin{cppcode}
std::vector<std::string> data{ "a", "b", "c", "d", "e", "f"};

std::copy(std::make_move_iterator(data.begin()), std::make_move_iterator(data.begin()+3), 
          data.begin()+3);
// data = { ?, ?, ?, "a", "b", "c" }
\end{cppcode}
\end{box-note}

When we need to copy to an overlapping range that isn’t suitable for forward copy, we can use the backward variant that copies in the opposite direction.

\cppversions{\texttt{copy\_backward}}{\CC98}{\CC20}{N/A}{\CC20}
\cppversions{\texttt{move\_backward}}{\CC11}{\CC20}{N/A}{\CC20}

\constraints{\texttt{bidirectional\_range -> bidirectional\_iterator}}{}{}{}

The output iterator cannot be within (first, last] and will be treated as the end iterator for the destination range, meaning that the algorithm will write the first value to std::prev(end).

\begin{box-note}
\begin{cppcode}
std::vector<std::string> data{ "a", "b", "c", "d", "e", "f"};

std::vector<std::string> out(9, "");
std::copy_backward(data.begin(), data.end(), out.end());
// "", "", "", "a", "b", "c", "d", "e", "f"

std::copy_backward(data.begin(), std::prev(data.end()), data.end());
// data = { "a", "a", "b", "c", "d", "e" }
\end{cppcode}
\end{box-note}

Similar to fill\_n and generate\_n, the standard also provides a copy\_n variant suitable for non-random-access ranges.

\cppversions{\texttt{copy\_n}}{\CC11}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{input\_iterator -> output\_iterator}}{\texttt{forward\_iterator -> forward\_iterator}}{}{}

The algorithm cannot check whether the requested count is valid and does not go out of bounds, so this burden is on the caller.

\begin{box-note}
\begin{cppcode}
std::list<int> data{1, 2, 3, 4, 5, 6, 7, 8, 9};
std::vector<int> out;

std::copy_n(data.begin(), 5, std::back_inserter(out));
// out == { 1, 2, 3, 4, 5 }
\end{cppcode}
\end{box-note}

\subsection{copy\_if, remove\_copy, remove\_copy\_if}

We can use copy\_if or remove\_copy\_if when we need to select only some of the elements to be copied.

\cppversions{\texttt{copy\_if}}{\CC11}{\CC20}{\CC17}{\CC20}
\cppversions{\texttt{remove\_copy, remove\_copy\_if}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{input\_range -> output\_range}}{\texttt{forward\_range -> forward\_iterator}}{N/A}{\texttt{unary\_predicate}}

These two algorithms provide the same logic, with opposite semantics: copy\_if will copy elements for which the predicate returns true, and remove\_copy\_if will copy elements for which the predicate returns false. Finally, remove\_copy copies elements that do not match the provided value.

\begin{box-note}
\begin{cppcode}
std::vector<int> data{ 1, 2, 3, 4, 5, 6, 7, 8, 9};
std::vector<int> even, odd, no_five;

auto is_even = [](int v) { return v % 2 == 0; };

std::ranges::copy_if(data, std::back_inserter(even), is_even);
// even == { 2, 4, 6, 8 }

std::ranges::remove_copy_if(data, std::back_inserter(odd), is_even);
// odd == { 1, 3, 5, 7, 9 }

std::ranges::remove_copy(data, std::back_inserter(no_five), 5);
// no_five == { 1, 2, 3, 4, 6, 7, 8, 9 }
\end{cppcode}
\end{box-note}

\subsection{sample}

A different twist on a selective copy is the sample algorithm.

\cppversions{\texttt{sample}}{\CC17}{N/A}{N/A}{\CC20}

\constraints{\texttt{forward\_range -> output\_iterator}}{}{}{}
\constraints{\texttt{input\_range -> random\_access\_iterator}}{}{}{}

The sample algorithm will copy a random selection of N elements from the source range to the destination range utilising the provided random number generator.

\begin{box-note}
\begin{cppcode}
std::vector<int> data{1, 2, 3, 4, 5, 6, 7, 8, 9};
std::vector<int> out;

std::sample(data.begin(), data.end(), std::back_inserter(out),
            5, std::mt19937{std::random_device{}()});
// e.g. out == {2, 3, 4, 5, 9} (guaranteed ascending, because source range is ascending)
\end{cppcode}
\end{box-note}

The two domains of this algorithm are due to the stable nature of the sampling, maintaining the order of elements from the source range.

This feature requires either the input range to be at least a forward range or the destination range needs to be a random-access range.

\subsection{replace\_copy, replace\_copy\_if}

We discussed the replace algorithm in a previous article in the series. The copy variants work identical, replacing values matching a predicate or a provided value. However, the result is output to a destination range instead of the replacement applied in place.

\cppversions{\texttt{replace\_copy, replace\_copy\_if}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{input\_range -> output\_iterator}}{\texttt{forward\_range -> forward\_iterator}}{N/A}{\texttt{unary\_predicate}}

The replace\_copy\_if variant replaces elements for which the predicate returns true. The replace\_copy replaces elements that match the provided value.

\begin{box-note}
\begin{cppcode}
std::vector<int> data{1, 2, 3, 4, 5, 6, 7, 8, 9};
std::vector<int> out, odd;

std::ranges::replace_copy(data, std::back_inserter(out), 5, 10);
// out == { 1, 2, 3, 4, 10, 6, 7, 8, 9 }

auto even = [](int v) { return v % 2 == 0; };
std::ranges::replace_copy_if(data, std::back_inserter(odd), even, -1);
// odd == { 1, -1, 3, -1, 5, -1, 7, -1, 9 }
\end{cppcode}
\end{box-note}

\subsection{reverse\_copy, rotate\_copy}

The last two algorithms from the copiers category change the order of copied elements.

\cppversions{\texttt{reverse\_copy}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{bidirectional\_range -> output\_iterator}}{\texttt{bidirectional\_range -> forward\_iterator}}{}{}

Not to be confused with copy\_backwards, which maintains the original order of elements, reverse\_copy will reverse the order of elements copied.

\begin{box-note}
\begin{cppcode}
std::vector<int> data{1, 2, 3, 4, 5, 6, 7, 8, 9};
std::vector<int> out;

std::ranges::reverse_copy(data, std::back_inserter(out));
// out == { 9, 8, 7, 6, 5, 4, 3, 2, 1 }
\end{cppcode}
\end{box-note}

\cppversions{\texttt{rotate\_copy}}{\CC98}{\CC20}{\CC17}{\CC20}

\constraints{\texttt{(forward\_range, forward\_iterator) -> output\_iterator}}{\texttt{(forward\_range, forward\_iterator) -> forward\_iterator}}{}{}

Following the logic of the rotate algorithm, rotate\_copy will first copy the elements starting from the designated middle element and then follow with the remaining elements from the beginning of the range.

\begin{box-note}
\begin{cppcode}
std::vector<int> data{1, 2, 3, 4, 5, 6, 7, 8, 9};
std::vector<int> out;

std::ranges::rotate_copy(data, data.begin()+4, std::back_inserter(out));
// out == { 5, 6, 7, 8, 9, 1, 2, 3, 4 }
\end{cppcode}
\end{box-note}
